---
title: "Argos-Links: a potential habitat distribution model for woodland species in Cornwall."
subtitle: "Technical summary"
author: | 
  | Hugh A. Graham, Alan Puttock, Richard E. Brazier
  | *Centre for Resilience in Environment, Water and Waste (CREWW), Geography, College of Life and  Environmental Sciences, University of Exeter, Exeter, UK*
output:
  pdf_document: default
  html_document:
    df_print: paged
fontfamily: mathpazo
fontsize: 11pt
urlcolor: blue
header-includes:
  - \usepackage{fancyhdr}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage[normalem]{ulem}
  - \usepackage{makecell}
  - \usepackage{xcolor}
  - \usepackage{titling}
  - \pretitle{\begin{flushleft}\fontsize{18bp}{18bp}\selectfont}
  - \posttitle{\end{flushleft}}
  - \preauthor{\begin{flushleft}\fontsize{13bp}{13bp}\selectfont}
  - \postauthor{\end{flushleft}}
  - \fancypagestyle{firststyle}{}
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(kableExtra)
library(dplyr)
library(tibble)
```

Argos-Links is a framework for deriving potential habitat models for a range of 
different terrestrial woodland species for Cornwall. The name Argos-links was 
chosen because the model works by calculating the relative distances/links
between woodland (**Argos in Cornish**) patches.

The model is a generalised solution for estimating potential habitat areas, producing 
two spatial polygon layers for a given species, the viable habitat in the zone and the 
area that contains this viable habitat, limited by road boundaries.

One of the key factors influencing terrestrial mammal mobility in a contemporary
landscape is the road network. Roads are by no means a permanent barrier to animals
but they are a significant cause of mortality and do exert and strong control 
over animal migration and movement (Glista, et al., 2009). Therefore, in this work
we make the simplified assumption that A and B roads are a barrier to the movement 
of the considered species. This simplification means that the model cannot be 
used for certain applications. For example it would not be suitable
for predicting the movement of animals between different viable habitat zones. 

However, The primary intention of this model is to identify viable habitat areas with sufficient 
woodland extent that is not intersected by A or B roads. This provides a useful indicator of 
which regions offer the best opportunities for releasing a given species. Those zones
that contain a greater woodland area should be considered to have a greater potential for 
species recovery/release success. 

The only habitat type considered in this model is woodland. We define woodland habitat as 
continuous woody cover with a minimum area of 0.1 ha and a minimum canopy cover of 25%.

There were four key variables considered in this model and are described in the 
following tables: 


### Habitat/Woodland preference
This may be Either Deciduous, or coniferous in the model. A mixed category is 
included in Table 1 for clarity. However, it is clear that all species can establish
in either woodland type.


*Table 1*
```{r table1-hab-defs, echo=FALSE, fig.cap="Table 1"}
habitat_defs <-
  tibble(
    Species = c('Boar',
                'Squirrel',
                'Marten',
                'Wildcat'),
    `Habitat Preference` = c(
      'Deciduous (1), Mixed (2), Coniferous (3) (Leaper et al. 1999; Łabudzki et al. 2009)',
      paste0(
        "Coniferous, Mixed, Deciduous (Lurz et al. 1995; Cagnin et al. 2000;",
        "Hämäläinen et al. 2018) but preference isn’t particularly strong. ",
        "However if considered in context of martens, coniferous habitat ",
        "likely less preferable when the latter present (Twining et al. 2022)."
      ),
      paste0(
        "No clear preference (Balharry et al. 2008), however broadleaved woodland ",
        "likely offers greater dietary breadth (Twining et al. 2022). "
      ),
      "Deciduous, Mixed, Coniferous (Sarmento et al. 2006; Oliveria et al. 2018)"
    )
  )
  
  kableExtra::kbl(habitat_defs, longtable = T, booktabs = T,align = c("c", "l")) %>%
    kable_styling(full_width = F, 
                  latex_options = c("repeat_header"),
                  font_size = 9) |>
    column_spec(1, width = "10em") |>
    column_spec(2, width = "30em")
```


### Minimum Total Woodland Area
 The minimum total accessible woodland area required for a given species. This 
 measure includes all woodland patches that are within the specified gap distance (Table 4)
 of one another.


*Table 2*
```{r table2-tot-wood-area, echo=FALSE}

select(habitat_defs, Species) %>%
  mutate(
    `tot. woodland area (ha)` = c(100, 1, 20, 100),
    `info` = c(
      paste0(
        "Mean home range size 4.9km2/2.5km2 for males/females in ",
        "Belgium (Prévot & Licoppe 2013), minimum size 1.06km2 in ",
        "Romania with mean daily home range between 0.6 – 1.25km2 ",
        "(Fodor et al. 2018). An adult female social unit of 6-30 ",
        "individuals may cover a range of 2-5km2 expanding to ",
        "10-40km2 in winter while adult male range can be 50km2. ",
        "(Spitz 1992). Density in mosaic landscape of France of 1-3 ",
        "animals per km2. Suggested home min home range 1km2."
      ),
      paste0(
        "Min home range 1.52ha in large Belgian forests ",
        "(Wauters & Dhondt 1990), densities of 0.5-1.5 per ha with large ",
        "fluctuations recorded (Lurz et al. 2005). Suggested min home ",
        "range 1 ha."
      ),
      paste0(
        "Mean home range in Bialoweiza Forest 2.23 km2 for males ",
        "and 1.49km2 for females (Zalewski et al. 2004), in more fragmented ",
        "habitat in the French Ardennes Minimum home range 0.25km2/0.10km2  ",
        "(mean 1.90/0.49) for males/females in fragmented habitat and ",
        "3.45km2/1.21km2 (mean 3.45/1.21) in forest (Mergey et al. 2011), ",
        "translocated martens in Wales mean 9.5km2 (range 0.2 -65.6km2) ",
        "(McNichol 2020). Mostly coniferous forest in Wales and mixed ",
        "woodland with agriculture in Ardennes, food availability likely ",
        "greatest influence over smaller home ranges sizes here and ",
        "especially within fragmented landscapes. Suggested min home ",
        "range 0.2km2."
      ),
      paste0(
        "In wet woodlands in Germany minimum home ranges of 8ha for ",
        "females and 30 ha for males, in Swiss mixed woodland surrounded by ",
        "pasture and woodland patches 70 ha for females and 150 ha for males ",
        "(Walsh 2019). German woodlands surrounded by cultural landscape host ",
        "home ranges averaging 285 ha for females and 1189 ha for males ",
        "(Jerosch et al. 2017). Suggested min home range 100ha."
      )
      
    )
  ) %>% 
  
  kbl(longtable = T, booktabs = T,align = c("c", "c" ,"l")) %>%
    kable_styling(full_width = F, 
                  latex_options = c("repeat_header"),
                  font_size = 9) |>
    column_spec(1, width = "5em") |>
    column_spec(2, width = "8em") |>
    column_spec(3, width = "30em")
```


### Minimum Patch Area
The minimum woodland patch size that might contribute to the habitat of a given 
species. 


*Table 3*
```{r table3-min-patch, echo=FALSE}
select(habitat_defs, Species) %>%
  mutate(
    `min. patch area (ha)` = c(25, 0.5, 20, 7),
    `info` = c(
      paste0(
        "Estimated 1km2 of woodland can sustainably support 4 boar in context ",
        "of Scottish reintroduction (Howells & Edward-Jones 1997) while taking ",
        "a less conservative estimate Leaper et al. (1999) suggested based on ",
        "various European populations an expected density of 3-5 boar per km2. ",
        "Suggested minimum patch size 0.25km2."
      ),
      paste0(
        "Minimum patch size on Isle of Wight 0.25 ha (Rushton et al. 1999). ",
        "Suggested minimum patch size 0.5ha"
      ),
      paste0(
        "In lowland linear riparian woods in Italy (similar habitat to Cornwall ",
        "– wooded river valleys below agri landscapes) estimated density of ",
        "pine martens is 0.8 - 2 per km2 (Pereboom et al 2008). Suggested ",
        "minimum patch size 0.2km2."
      ),
      paste0(
        "Minimum individual core range within wider home range 7ha for females ",
        "and 215ha for males (433/770 max) in woodland/cultural landscape in ",
        "Germany (Jerosch et al. 2017)."
      )
    ))%>% 
  
  kbl(longtable = T, booktabs = T,align = c("c", "c" ,"l")) %>%
    kable_styling(full_width = F, 
                  latex_options = c("repeat_header"),
                  font_size = 9) |>
    column_spec(1, width = "5em") |>
    column_spec(2, width = "8em") |>
    column_spec(3, width = "30em")
```


### Gap distance
The distance between individual woodland patches that is likely to be travelled 
by a species within its home range.


*Table 4*
```{r table4-gap-dist, echo=FALSE}
select(habitat_defs, Species) %>%
  mutate(
    `max. gap distance (m)` = c(2980, 51, 1100, 1430),
    `info` = c(
      paste0(
        "Mean dispersal distance 2.98km in Belguim (Prévot & Licoppe 2013), ",
        "direct distance between resting places varies between 0-7km ",
        "(Spitz 1986). "),
      paste0(
        "Minimum dispersal distance 51m in Belgium in fragmented habitats ",
        "(mean 1,014m) (Wauters et al 2010)"),
      paste0(
        "Minimum dispersal distance of translocated martens in Wales was 1.1km ",
        "(mean 8.7km) (McNichol et al. 2020)."),
      paste0(
        "Unknown. Maximum mean distance moved by wildcats in a Portugese study ",
        "was 1.43km (Matias et al. 2021). ")
      )
    )%>% 
  
  kbl(longtable = T, booktabs = T,align = c("c", "c" ,"l")) %>%
    kable_styling(full_width = F, 
                  latex_options = c("repeat_header"),
                  font_size = 9) |>
    column_spec(1, width = "5em") |>
    column_spec(2, width = "8em") |>
    column_spec(3, width = "30em")
```


## Methods

The Argos-Links models is written using the R programming language and is dependent
on the following packages: "sf", "plyr", "tidyverse", "purrr", "furrr", "curl", "zip", 
"terra", "raster", "fasterize", "gdalio", "rmapshaper", "stars", "tmap".

The model is written as a reproducible {targets} workflow. The source code for 
the model can be found at: [https://github.com/h-a-graham/argoslinks](https://github.com/h-a-graham/argoslinks)

### Input Data
The following input datasets are used for this model and should be additionally
sited when referring to this model. See table X for the dataset names and references.

```{r input-data-table, echo=FALSE}

tibble(
  `Full Name` = c("Centre for Ecology and Hydrology, Land Cover Map 2019",
                  "Copernicus Tree Cover Density, 2018", "Ordnance Survey Boundary Line",
                  "Ordnance Survey Open Roads"),
  
  `Short Name` = c("LCM 2019", "TCD 2018",
           "OS Boundary-Line", "OS open-roads"),
  `source location` = 
  c("https://catalogue.ceh.ac.uk/documents/643eb5a9-9707-4fbb-ae76-e8e53271d1a0",
    "https://land.copernicus.eu/pan-european/high-resolution-layers/forests/tree-cover-density/status-maps/tree-cover-density-2018",
    "https://www.ordnancesurvey.co.uk/business-government/products/boundaryline",
    "https://www.ordnancesurvey.co.uk/business-government/products/open-map-roads"
     )
) %>% 
  # mutate(`source location` = cell_spec(`source location`, "html", link = `source location`)) %>%
  # kable(escape = FALSE) %>%
  kbl(longtable = T, booktabs = T,align = c("c", "c" ,"l")) %>%
    kable_styling(full_width = F, 
                  latex_options = c("repeat_header"),
                  font_size = 9) %>% 
    column_spec(1, width = "8em") |>
    column_spec(2, width = "5em") |>
    column_spec(3, width = "30em")

```

### Data Preparation:

- Cornwall county region from the Ordnance Survey Boundary-Lines dataset Extracted
as a spatal polygon feature.

- CEH Land Cover Map and Copernicus Tree Cover were Reprojected to 10 m resolution, with 
equivalent extents and grid alignment, contrained by the cornwall county bounding 
extent. The LCM was resampled using the "nearest" resampling method and the TCD is
resampled using the "cubic-spline" method. This operation is handled using the 
gdal-warp utility.


- A new raster is reated from the LCM and TCD maps. Regions where the TCD value 
is >25 were given a value of 1. Regions where the LCM is classed as Broadleaf 
woodland were given a value of 1. Regions where the LCM is classed as coniferous 
were given a value of 2 (overriding a value of 1 if previously assigned.) Values
not considered woodland were set to 0. Offshore regions were set as NA. We refer 
to this raster as the Woodland Type Raster (WTR)


- All Roads that intersect the Area of Interest (AOI), in this case the Cornwall
county boundary, were extracted from the OS open-roads dataset. 

### Creating woodland spatial vector

- WTR values that equaled zero were converted to NA. The WTR is then converted to a 
spatial vector where broadleafed and coniferous were each classified with a 
unique integer ID (either 1 or 2). The features that fall within 1 km of the
county boundary were retained.

- Woodland polygons with an area < 0.1 ha were removed, as these were not 
considered to reliably contribute valuable woodland habitat. 

- The feature geometries were then simplified, retaining 25% of vertices. This 
improves the performance of spatial operation whilst retaining accuracy.

### splitting woodland spatial vector by road network.

- The cornish road network was filtered, selecting only A and B roads. The 
network was then unionised into a single MULTILINESTRING geometry. Herein 
referred to as minimal road network.

- A negative buffer of 500 m was applied to the county area and was then split
into discrete polygons using minimal road network. The - 500 m buffer is used
so that small gaps between road dead-ends and the coastline were not considered
as viable connectivity routes. This simplification is required in order to avoid
the assumption that viable corridors can form along these areas. In reality, 
they typically comprise steep cliffs/rocky areas/beaches which were less likely to 
afford a viable route between woodland patches. The polygon areas created in this
step are referred to as road patches.

- Woodland areas that intersect the road patches were retained and a spatial 
intersection of the road patches and woodland areas is carried out, resluting in
a woodland area polygon split by the minimal road network. These features are 
referred to herein as in-patch woodland.

- The previous intersection step excludes woodland that falls within 500 m of the 
coastline. These woodlands may provide important habitat. Therefore, woodland 
polygons that either overlap or do not intersect the road patches were located and subset.
Features of this subset that overlap the road pataches are intersected (removing the part of
the polygon that is already considered). These features are then then spatially
joined to the nearest road patch using a k-nearest neighbours algorithm with
a maximum search distance of 750 m. These features are referred to herein as 
out-patch woodland.

- The in-patch and out-patch woodland features are joined resulting in a woodland
area polygon with unique ID attributes, a road patch attribute which denotes the
inter-road zone that each woodland polygon belongs to and a woodland type classifier.

- This spatial polygon layer is provided as a layer in the provided dataset (Named:
MW_InterRoadCover). This dataset is not species specific but is general and provides
a useful overview as to which regions of the county contain woodland (not) intersected 
by main roads.

### Species-level models.

- At this stage, the rules described in Tables 2-4 are used to determine which
inter-road zones provide suitable habitat. 

- Woodland features are grouped by the road patch ID. For each road patch the 
following steps are carried out: 

  - The area of each invidual woodland patch is calculated. All features with an
  area less than the minimum patch size (Table 3) are removed.
  
  - The total area of remaining woodland within the road patch is calculated. If this is
  less than the minimum Total Woodland Area (Table 2) for the species, all
  woodland within the roach patch is removed and no features are returned. 
  
  - Remaining feature are all buffered by the gap distance (Table 4) divided by 2.
  The distance is halved because other woodlands are also buffered and this therefore
  marks the halfway point between woodland areas. Where these buffers intersect,
  the gap distance is less than the maximum allowed.
  
  - These buffered areas are then intersected by remaining woodland greater than
  the minimum patch size. Those intersectng areas are retained as viable habitat. 
  
  - The Total habitat area is once again checked and if less than the minimum 
  total area, no features are returned.

- Returned woodland features and and buffered areas were retained and spatially
joined with the road patches to assign a unique road patch ID for each zone and 
the patches within the zones. 

- These features were then saved in the GeoPackage format.


### Output Data

The model results are provided in the directory ArgosLinks/Data/. Data is provided
in the GeoPackage format which supports multiple layers and can be read my the 
majority of GIS software. Two files are provided: *ArgosLinks.gpkg* and 
*MW_InterRoadCover.gpkg*.


#### *ArgosLinks.gpkg*

The file contains 8 layers with the one of the follwoing naming conventions: 

ArgosLinks_{SPECIES}_patches or ArgosLinks_{SPECIES}_chunks. 

layers ending with *patches* display the woodland area suitable for a given species. 
Those ending in *chunks* display the area accessible by the species based on range 
and road obstructions. Both patch and chunk laers have corresponding **species_region**
and **inter_road_zone** attributes that identify habitat regions. 

#### *MW_InterRoadCover.gpkg*

This file comprises a single layer displaying all woodland zones. All zones have
a **inter_road_zone** attribute that corresponds with the ArgosLinks dataset. 
The woodland type - either Broadleaf or Confier is also given as **wood_class**. 


### Output Interactive Maps

interactive leafleat maps were produced to display the general datasets and 
individual species level models. And can be found in the following directory: 
ArgosLinks/Data/LeafMaps. These html files can be viewed in any web browser.



\pagebreak
## References

Balharry, E. (1993) Social organization in martens: an inflexible system? 
Symposia of the Zoological Society of London, 65, 321-345.

Balharry, E., Jefferies, D. J., Birks, J. D. S. (2008) Pine marten. In the 
Handbook of British Mammals 4th Edition, pp.447-455. Edited by Hariss, S.. & 
Yalden, D. W. The Mammal Society, Southampton.

Brainerd, S.M., Helldin, J.O., Lindström, E.R., Rolstad, E., Rolstad, J. & 
Storch, I., (1995) Pine marten (Martes martes) selection of resting and 
denning sites in Scandinavian managed forests. In Annales Zoologici Fennici 
(pp. 151-157). Finnish Zoological and Botanical Publishing Board.

Cagnin, M., Aloise, G., Fiore, F., Oriolo, V., & Wauters, L. A. (2000) Habitat 
use and population density of the red squirrel, Sciurus vulgaris meridionalis, 
in the Sila Grande mountain range (Calabria, South Italy). Italian Journal of 
Zoology, 67(1), 81-87.

Glista, D. J., DeVault, T. L., & DeWoody, J. A. (2009). A review of mitigation 
measures for reducing wildlife mortality on roadways. Landscape and urban 
planning, 91(1), 1-7.

Hämäläinen, S., Fey, K., & Selonen, V. (2018) Habitat and nest use during natal 
dispersal of the urban red squirrel (Sciurus vulgaris). Landscape and Urban 
Planning, 169, 269-275.

Howells, O. and Edwards-Jones, G. (1997). A feasibility study of reintroducing 
wild boar Sus scrofa to Scotland: Are existing woodlands large enough to support 
minimum viable populations. Biological Conservation, 81, 77-89.

Jerosch, S., Götz, M. and Roth, M., (2017) Spatial organisation of European 
wildcats (Felis silvestris silvestris) in an agriculturally dominated landscape 
in Central Europe. Mammalian Biology, 82, pp.8-16.

Fodor, J.T., (2018) Wild Boar 2018. Home Range and Habitat Use in Two Romanian 
Habitats. Acta Silvatica et Lignaria Hungarica, 14(1), pp.51-63.

Łabudzki, L., Górecki, G., Skubis, J. and Wlazełko, M., (2009) Forest habitats 
use by wild boar in the Zielonka Game Investigation Centre. Acta Scientiarum 
Polonorum Silvarum Colendarum Ratio et Industria Lignaria, 8(4), pp.51-57.

Leaper, R., Massei, G., Gorman, M. and Aspinall, R. (1999) The feasibility of 
reintroducing Wild Boar (Sus scrofa) to Scotland.

Lurz, P. W., Garson, P. J., & Rushton, S. P. (1995) The ecology of squirrels in 
spruce dominated plantations: implications for forest management. Forest ecology 
and management, 79(1-2), 79-90.

Lurz, P. W. W., Gurnell J., and Magris, L. (2005) Sciurus vulgaris. Mammalian 
Species 769: 1-10.

Mergey, M., Helder, R. and Roeder, J.J., (2011) Effect of forest fragmentation on 
space-use patterns in the European pine marten (Martes martes). Journal of 
Mammalogy, 92(2), pp.328-335.

Oliveira, T., Urra, F., López‐Martín, J.M., Ballesteros‐Duperón, E., Barea‐Azcón,
J.M., Moléon, M., Gil‐Sánchez, J.M., Alves, P.C., Díaz‐Ruíz, F., Ferreras, P. and
Monterroso, P., (2018) Females know better: Sex‐biased habitat selection by the
European wildcat. Ecology and evolution, 8(18), pp.9464-9477.

Pereboom, V., Mergey, M., Villerette, N., Helder, R., Gerard, J.F. and Lode, T., 
(2008) Movement patterns, habitat selection, and corridor use of a typical
woodland-dweller species, the European pine marten (Martes martes), in fragmented landscape. Canadian Journal of Zoology, 86(9), pp.983-991.

Rushton, S.P., Lurz, P.W.W., South, A.B. and Mitchell‐Jones, A., (1999)
Modelling the distribution of red squirrels (Sciurus vulgaris) on the Isle of 
Wight. Animal Conservation, 2(2), pp.111-120.

Sarmento, P., Cruz, J., Tarroso, P. and Fonseca, C., (2006) Space and habitat 
selection by female European wild cats (Felis silvestris silvestris). Wildlife 
Biology in Practice, pp.79-89.

Spitz, F. (1986) Current state of knowledge of wild boar biology. Pig News and 
Information, 7 (2), 171–175.

Spitz, F. (1992) General model of spatial and social organisa- tion of the wild 
boar (Sus scrofa). In Ongules/Ungulates 91, ed. F. Spitz, G. Janeau, G. Gonzalez 
and S. Aulangier, pp. 385-389. SFEPM-IRGM, Toulouse.

Twining, J. P., Sutherland, C., Reid, N., and Tosh, D. G. (2022) Habitat mediates 
coevolved but not novel species interactions. Proceedings of the Royal 
Society B, 289, 20212338.

Wauters, L.A. and Dhondt, A.A., (1990) Nest-use by red squirrels (Sciurus 
vulgaris Linnaeus, 1758).

Wauters, L.A., Verbeylen, G., Preatoni, D., Martinoli, A. and Matthysen, E., 
(2010) Dispersal and habitat cuing of Eurasian red squirrels in fragmented 
habitats. Popul Ecol, 52, pp.527-536.

Zalewski, A., Jedrzejewski, W. and Jedrzejewski, B., (2004) Mobility and home 
range use by pine martens (Martes martes) in a Polish primeval forest. 
Ecoscience, 11(1), pp.113-122.
