---
title: "Argos Links"
author: "Hugh Graham"
date: '2022-06-02'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(kableExtra)
library(dplyr)
library(tibble)
```

Argos-Links is a framework for deriving potential habitat models for a range of 
different terrestrial woodland species for Cornwall. The name Argos-links was 
chosen as a name because the model works by calculating the relative distances/links
between woodland (**Argos in cornish**) patches.

The model is a generalised solution for estimating potential habitat areas, producing 
two spatial polygon layers for a given species, the viable habitat in the zone and the 
area that contains this viable habitat (the convex hull of the combined are of the patches).

One of the key factors influencing terrestrial mammal mobility in a contemporary
landscape is the road network. Roads are by no means a permenant barrier to animals
but they are a significant cause of mortality and do exert and strong control 
over the animal migration and movemet. Therefore, A and B roads are considered 
as a barrier to the movement of a species. This simplification means that the 
model cannot be used for certain applications. For example it would not be suitable
for predicting the movement of animals between different viable habitat zones. 

The primary intention of this model is to identify viable habitat areas with sufficient 
woodland extent that is not intersecte by A or B roads. This provides a useful indicator of 
which regions offer the best opportunities for releasing a given species. Those zones
that contain a greater woodland area should be considered to have a greater potential for 
species success. 

There are four key variables considered in this model: 
woodland type:
gap distance:
zone size:
patch min:

This Documetn summarises the methodology 

### Habitat/Woodland preference
```{r cars, echo=FALSE}
habitat_defs <-
  tibble(
    Species = c('Boar',
                'Squirrel',
                'Marten',
                'Wildcat'),
    `Habitat Preference` = c(
      'Deciduous (1), Mixed (2), Coniferous (3) (Leaper et al. 1999; Łabudzki et al. 2009)',
      paste0(
        "Coniferous, Mixed, Deciduous (Lurz et al. 1995; Cagnin et al. 2000;",
        "Hämäläinen et al. 2018) but preference isn’t particularly strong. ",
        "However if considered in context of martens, coniferous habitat ",
        "likely less preferable when the latter present (Twining et al. 2022)."
      ),
      paste0(
        "No clear preference (Balharry et al. 2008), however broadleaved woodland ",
        "likely offers greater dietary breadth (Twining et al. 2022). "
      ),
      "Deciduous, Mixed, Coniferous (Sarmento et al. 2006; Oliveria et al. 2018)"
    )
  )
  
  kableExtra::kbl(habitat_defs, longtable = T, booktabs = T,align = c("c", "l")) %>%
    kable_styling(full_width = F, 
                  latex_options = c("repeat_header"),
                  font_size = 9) |>
    column_spec(1, width = "10em") |>
    column_spec(2, width = "30em")
```



### Minimum Total Woodland Area
```{r tot-wood-area, echo=FALSE}

select(habitat_defs, Species) %>%
  mutate(
    `tot. woodland area (ha)` = c(100, 1, 20, 100),
    `info` = c(
      paste0(
        "Mean home range size 4.9km2/2.5km2 for males/females in ",
        "Belgium (Prévot & Licoppe 2013), minimum size 1.06km2 in ",
        "Romania with mean daily home range between 0.6 – 1.25km2 ",
        "(Fodor et al. 2018). An adult female social unit of 6-30 ",
        "individuals may cover a range of 2-5km2 expanding to ",
        "10-40km2 in winter while adult male range can be 50km2. ",
        "(Spitz 1992). Density in mosaic landscape of France of 1-3 ",
        "animals per km2. Suggested home min home range 1km2."
      ),
      paste0(
        "Min home range 1.52ha in large Belgian forests ",
        "(Wauters & Dhondt 1990), densities of 0.5-1.5 per ha with large ",
        "fluctuations recorded (Lurz et al. 2005). Suggested min home ",
        "range 1 ha."
      ),
      paste0(
        "Mean home range in Bialoweiza Forest 2.23 km2 for males ",
        "and 1.49km2 for females (Zalewski et al. 2004), in more fragmented ",
        "habitat in the French Ardennes Minimum home range 0.25km2/0.10km2  ",
        "(mean 1.90/0.49) for males/females in fragmented habitat and ",
        "3.45km2/1.21km2 (mean 3.45/1.21) in forest (Mergey et al. 2011), ",
        "translocated martens in Wales mean 9.5km2 (range 0.2 -65.6km2) ",
        "(McNichol 2020). Mostly coniferous forest in Wales and mixed ",
        "woodland with agriculture in Ardennes, food availability likely ",
        "greatest influence over smaller home ranges sizes here and ",
        "especially within fragmented landscapes. Suggested min home ",
        "range 0.2km2."
      ),
      paste0(
        "In wet woodlands in Germany minimum home ranges of 8ha for ",
        "females and 30 ha for males, in Swiss mixed woodland surrounded by ",
        "pasture and woodland patches 70 ha for females and 150 ha for males ",
        "(Walsh 2019). German woodlands surrounded by cultural landscape host ",
        "home ranges averaging 285 ha for females and 1189 ha for males ",
        "(Jerosch et al. 2017). Suggested min home range 100ha."
      )
      
    )
  ) %>% 
  
  kbl(longtable = T, booktabs = T,align = c("c", "c" ,"l")) %>%
    kable_styling(full_width = F, 
                  latex_options = c("repeat_header"),
                  font_size = 9) |>
    column_spec(1, width = "10em") |>
    column_spec(2, width = "10em")|>
    column_spec(2, width = "25em")
```

### Minimum Patch Size

```{r min-patch, echo=FALSE}
select(habitat_defs, Species) %>%
  mutate(
    `min. patch area (ha)` = c(25, 0.5, 20, 7),
    `info` = c(
      paste0(
        "Estimated 1km2 of woodland can sustainably support 4 boar in context ",
        "of Scottish reintroduction (Howells & Edward-Jones 1997) while taking ",
        "a less conservative estimate Leaper et al. (1999) suggested based on ",
        "various European populations an expected density of 3-5 boar per km2. ",
        "Suggested minimum patch size 0.25km2."
      ),
      paste0(
        "Minimum patch size on Isle of Wight 0.25 ha (Rushton et al. 1999). ",
        "Suggested minimum patch size 0.5ha"
      ),
      paste0(
        "In lowland linear riparian woods in Italy (similar habitat to Cornwall ",
        "– wooded river valleys below agri landscapes) estimated density of ",
        "pine martens is 0.8 - 2 per km2 (Pereboom et al 2008). Suggested ",
        "minimum patch size 0.2km2."
      ),
      paste0(
        "Minimum individual core range within wider home range 7ha for females ",
        "and 215ha for males (433/770 max) in woodland/cultural landscape in ",
        "Germany (Jerosch et al. 2017)."
      )
    ))%>% 
  
  kbl(longtable = T, booktabs = T,align = c("c", "c" ,"l")) %>%
    kable_styling(full_width = F, 
                  latex_options = c("repeat_header"),
                  font_size = 9) |>
    column_spec(1, width = "10em") |>
    column_spec(2, width = "10em")|>
    column_spec(2, width = "25em")
```

### Gap distance

```{r gap-dist, echo=FALSE}
select(habitat_defs, Species) %>%
  mutate(
    `max. gap distance (m)` = c(2980, 51, 1100, 1430),
    `info` = c(
      paste0(
        "Mean dispersal distance 2.98km in Belguim (Prévot & Licoppe 2013), ",
        "direct distance between resting places varies between 0-7km ",
        "(Spitz 1986). "),
      paste0(
        "Minimum dispersal distance 51m in Belgium in fragmented habitats ",
        "(mean 1,014m) (Wauters et al 2010)"),
      paste0(
        "Minimum dispersal distance of translocated martens in Wales was 1.1km ",
        "(mean 8.7km) (McNichol et al. 2020)."),
      paste0(
        "Unknown. Maximum mean distance moved by wildcats in a Portugese study ",
        "was 1.43km (Matias et al. 2021). ")
      )
    )%>% 
  
  kbl(longtable = T, booktabs = T,align = c("c", "c" ,"l")) %>%
    kable_styling(full_width = F, 
                  latex_options = c("repeat_header"),
                  font_size = 9) |>
    column_spec(1, width = "10em") |>
    column_spec(2, width = "10em")|>
    column_spec(2, width = "20em")
```


## Methods

The Argos-Links models is written using the R programming language and is dependent
on the following packages: ...

The model is written as a reproducible {targets} workflow. The source code for 
the model can be found here: [LINK TO CODE]

### Input Data
THe following input datasets are used for this model and should be additionally
sited when referring to this model. See table X for the dataset names and references.

```{r input-data-table, echo=FALSE}

tibble(
  Name = c("CEH LCM 2019", "Copernicus TCD 2018", 
           "National Forest Inventory 2018", "OS vectorMap District"),
  `source location` = 
  c("https://catalogue.ceh.ac.uk/documents/643eb5a9-9707-4fbb-ae76-e8e53271d1a0",
    "https://land.copernicus.eu/pan-european/high-resolution-layers/forests/tree-cover-density/status-maps/tree-cover-density-2018",
    "https://data-forestry.opendata.arcgis.com/datasets/d3d7bfba1cba4a3b83a948f33c5777c0_0/explore",
    " https://osdatahub.os.uk/downloads/open/VectorMapDistrict"
     )
) %>% 
  mutate(`source location` = cell_spec(`source location`, "html", link = `source location`)) %>%
  kable("html", escape = FALSE) %>%
  kable_styling(bootstrap_options = c("hover", "condensed"),
                full_width = F, 
                  latex_options = c("repeat_header"),
                  font_size = 9) %>% 
    column_spec(1, width = "10em") %>% 
    column_spec(2, width = "300em") 

```


### spatial workflow:




## References
