---
title: "Argos Links"
author: "Hugh Graham"
date: '2022-06-02'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(kableExtra)
library(dplyr)
```

Argos-Links is a framework for deriving potential habitat models for a range of 
different terrestrial woodland species for Cornwall. The name Argos-links was 
chosen as a name because the model works by calculating the relative distances/links
between woodland (**Argos in cornish**) patches.

The model is a generalised solution for estimating potential habitat areas, producing 
two spatial polygon layers for a given species, the viable habitat in the zone and the 
area that contains this viable habitat (the convex hull of the combined are of the patches).

One of the key factors influencing terrestrial mammal mobility in a contemporary
landscape is the road network. Roads are by no means a permenant barrier to animals
but they are a significant cause of mortality and do exert and strong control 
over the animal migration and movemet. Therefore, A and B roads are considered 
as a barrier to the movement of a species. This simplification means that the 
model cannot be used for certain applications. For example it would not be suitable
for predicting the movement of animals between different viable habitat zones. 

The primary intention of this model is to identify viable habitat areas with sufficient 
woodland extent that is not intersecte by A or B roads. This provides a useful indicator of 
which regions offer the best opportunities for releasing a given species. Those zones
that contain a greater woodland area should be considered to have a greater potential for 
species success. 

There are four key variables considered in this model: 
woodland type:
gap distance:
zone size:
patch min:

This Documetn summarises the methodology 

### Habitat/Woodland preference
```{r cars, echo=FALSE}
habitat_defs <-
  tibble::tibble(
    Species = c('Boar',
                'Squirrel',
                'Marten',
                'WildCat'),
    `Habitat Preference` = c(
      'Deciduous (1), Mixed (2), Coniferous (3) (Leaper et al. 1999; Łabudzki et al. 2009)',
      paste0(
        "Coniferous, Mixed, Deciduous (Lurz et al. 1995; Cagnin et al. 2000;",
        "Hämäläinen et al. 2018) but preference isn’t particularly strong. ",
        "However if considered in context of martens, coniferous habitat ",
        "likely less preferable when the latter present (Twining et al. 2022)."
      ),
      paste0(
        "No clear preference (Balharry et al. 2008), however broadleaved woodland ",
        "likely offers greater dietary breadth (Twining et al. 2022). "
      ),
      "Deciduous, Mixed, Coniferous (Sarmento et al. 2006; Oliveria et al. 2018)"
    )
  )
  
  kableExtra::kbl(habitat_defs, longtable = T, booktabs = T,align = "c") %>%
    kable_styling(full_width = F, 
                  latex_options = c("repeat_header"),
                  font_size = 9) |>
    column_spec(1, width = "10em") |>
    column_spec(2, width = "30em")
```


### Minimum Patch Size

```{r}

```

### Minimum Total Woodland Area

```{r}

```

### Gap distance

```{r}

```


## Methods

The Argos-Links models is written using the R programming language and is dependent
on the following packages: ...

The model is written as a reproducible {targets} workflow. The source code for 
the model can be found here: [LINK TO CODE]

### Input Data
THe following input datasets are used for this model and should be additionally
sited when referring to this model. See table X for the dataset names and references.

```{r input-data-table, echo=FALSE}

```


### spatial workflow:




## References
